Privacy-Respecting Type Error Telemetry at Scale

### 2023-09-30 programming23/paper.tex
related work todos

% TODO BEFORE SUBMISSION
% be consistent between "IDs" and "ids" [SK prefers "IDs" to avoid it being read as a word]
% even more so between "ID" and "id" [NB: "id" is a word in English]
% \usepackage[sort]{natbib} or otherwise manually fix cites to be in ascending order
% redo fig 5 to have better resolution

@@

Creators program using the
{Luau} programming language~\cite{luau-lang.org},
an extension of {Lua~5.1~\cite{lua}}.
The main addition in Luau is a static type system that infers
types for all Luau programs on the fly, as creators modify the code.
These types are used primarily in IDE tooling such as autocomplete and
in API documentation~\cite{luau-autocomplete}, and indeed, creators
may be unaware that a typechecker is analyzing their code.
However, creators can opt in to receiving type error reports and they can write
their own types to guide designs and document their intentions.

Due to the broad community of creators, the goals of the Luau type
system are rather unique~\cite{bfj-hatra-2021}.
Whereas a traditional type system focuses on compilation and memory
safety, Luau takes a lenient approach by default and lets creators
gradually~\cite{st-sfp-2006,tf-popl-2008} migrate to rigorous checks
one module at a time.
In this way, for example,
students in code camps may begin in the \mnocheck{} mode that reports only
syntax errors; hobbyists can use \mnonstrict{} mode to identify
would-be runtime errors; and developers who need to ensure software
quality can rely on \mstrict{} mode.
(No matter the mode, a {\FS{}} type analysis still drives the IDE tools.)
Luau's types aim to support untyped designs, in the spirit of migratory
typing~\cite{tfffgksst-snapl-2017}, so that creators can switch modes without
needing to restructure their code in major, potentially-breaking, ways.

@@

{Roblox Studio} is a workbench that
combines {3D creation} tools and an Integrated
Developer Environment (IDE), as seen in \cref{fig:roblox-studio}.
The IDE includes an optional \emph{Script Analysis} widget that
reports syntax errors, type errors, and problems identified by
lint tools. The main editor widget can also highlight
source locations in code where reported errors occur.

@@

\begin{figure}[t]\centering
  \includegraphics[width=.35\textwidth]{img/roblox-studio.png}
  \hspace{1cm}
  \includegraphics[width=.35\textwidth]{img/roblox-studio-ide.png}
  \caption{{Roblox Studio 3D creation} tools (left) and IDE (right).}
  \label{fig:roblox-studio}
\end{figure}

@@

In addition to the main type analysis, Roblox Studio takes a second pass over
every codebase with a \emph{\FS{}} analysis to infer autocomplete suggestions
and drive other IDE tools.
To infer precise types,
the \FS{} analysis uses rigorous checks very similar to those of
\mstrict{} mode (but not identical: see \cref{s:strict-vs-forcedstrict})
no matter what mode the creator declared for the script.
(Internally to Roblox Studio, this analysis is called \emph{forced strict}
because it ignores the creator's choice and applies strong checks.)
When it fails to infer a type, \FS{} analysis defaults to the unknown gradual
type~\cite{st-sfp-2006}.
Creators cannot see the errors reported by \FS{} analysis.

@@

Roblox Studio has a framework for reporting telemetry.
Various features in the IDE use telemetry to measure their effectiveness.
The Luau open source toolchain does \emph{not} report telemetry,
as it is designed to be used in build environments such as Continuous Integration
servers, where hermetic deterministic builds are expected.

@@

\begin{figure}[t]\centering
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline
    ID & Time & Mode & Reason & Size Info & Overall Counts & Edit Range Counts
    \\\toprule
    \multicolumn{6}{c}{\raisebox{.5\normalbaselineskip}[0pt][0pt]{$\underbrace{\hspace{9cm}}$}} &
    \multicolumn{1}{c}{\raisebox{.5\normalbaselineskip}[0pt][0pt]{$\underbrace{\hspace{3cm}}$}} \\[-0.5ex]
    \multicolumn{6}{c}{fixed-size} & \multicolumn{1}{c}{variable-size}
  \end{tabular}
  \caption{Structure of a telemetry record. There are 13 overall counts and up to 70 edit range counts in each record.}
  \label{f:tdata}
\end{figure}

@@

\Cref{f:tdata} shows the structure of our telemetry records:
\begin{enumerate}
  \item
    ID: a \panon{} identifier for the current session.
  \item
    Time: one timestamp from the client and one from the server.
    %% timestamp explanation \cref{s:data-cleaning}
  \item
    Mode: type analysis mode of the current file: \mnocheck{},
    \mnonstrict{}, or \mstrict{}.
  \item
    Reason: a flag that explains whether this record was sent due to a
    randomly-selected keystroke or a module switch.
  \item
    Size Info: number of lines in the codebase and number of lines in the edit range.
  \item
    Overall Counts: summarizes type errors and \FS{} errors during the
    last two invocations of type analysis, and additionally reports the number
    of times type analysis hit an internal limit on problem
    size~(see~\cref{s:code-too-complex} for details).
    For each invocation and each kind of
    analysis error, there are three summary counts:
    %% grand total = 12
    \begin{itemize}
      \item total number of errors,
      \item errors in the current module, and
      \item errors in the current edit range.
    \end{itemize}
  \item
    Edit Range Counts: these list specific errors that arose in the last two
    type analyses and overlap with the current edit range.
    Since there are 35 possible errors~(\cref{s:context}), there can be up to 70 counts
    in a telemetry record.
    Exactly how to interpret this data depends on which analysis it came from:
    \begin{itemize}
      \item
        Errors in the latest type analysis are currently visible to the
        creator.
        These may have been introduced by the changes in the edit range.
      \item
        Errors from the previous type analysis were visible before the creator
        made the latest edits.
        These may have motivated some of the changes in the edit range.
    \end{itemize}
\end{enumerate}

@@

Luau has an efficient type checker that can
analyze thousands of lines of code~(\cref{t:codebase-size}).
But, few sessions use type analysis directly and the data presents little
evidence that doing so would improve creators' quality of life.
We conjecture that Luau needs further tailoring to match
common idioms in Roblox scripts.
Discovering these idioms is a question for future work~(\cref{s:conclusion}).

@@

Third, session-based data has limitations.
A creator who closes and reopens Roblox Studio every day has
a much higher chance of generating telemetry than one who
leaves Studio open all week.
We have no way to reliably check whether multiple sessions are
reporting on the same codebase.
Sessions may end abruptly, for instance because of machine reboot,
instead of ending when the creator was finished editing code.

@@

Our telemetry has two distinguishing characteristics:
it includes no private information (both PII and source code details),
and it sends data on
randomly-selected runs of the type checker rather than specific events of interest.
To the best of our knowledge, this design occupies a unique position relative to prior
work~(\cref{t:telemetry-design}).\todo{Where do these labels come
  from?!? There are no ``prior work'' citations either!}

In other professional-grade IDEs, such as VSCode~\cite{vscode-telemetry},
IntelliJ~\cite{intellij-telemetry}, and \code{.NET}~\cite{dotnet-telemetry}, telemetry may include all sorts of
data: error reports, source code fragments, time\-stamps, and filepaths.
Users may be able to opt out of some telemetry, but the details depend on the
license agreement.
Furthermore, at least in VSCode, IDE extensions can report their own telemetry.
While this data can be invaluable for discovering bugs in production, it
must be handled with extreme care.

The Blackbox project similarly collects telemetry in broad strokes.
Every Java program submitted for compilation goes into the Blackbox database (minus
any comments above the main class) along with the compiler's response~\cite{bkmu-sigcse-2014}.
%% Though somewhat intrusive,
The millions of programs in the Blackbox dataset
have enabled dozens of contributions to the area of CS
education~\cite{bask-icer-2018}. However, these operate at the
opposite end of privacy.

@@

At the opposite end of the privacy spectrum, Go's design for \emph{transparent
telemetry} reports only counter values~\cite{transparent-telemetry}.
Unlike Roblox Studio, transparent telemetry includes no timestamps and no 
session ids.
While useful for learning about the frequency and distribution of
specific events, the lack of timestamps and ids means it cannot track
edits over time (which we use in \cref{f:error-density}).
% BEN: commented for space.
% it cannot track 
% The data is useful for learning about the frequency of specific events and the
% distribution of events across clients (e.g., how many \code{CodeTooComplex}
% errors arise? what is the average number of type errors?).
% However, the lack of timestamps and ids means that it cannot track edits
% over time (a feature that is quite useful to us, for example, in~\cref{f:error-density}).
% In~\cref{f:error-density}, for example, we see that increases to the number
% of type errors are quickly followed by decreases.
% Without timestamps, we would be unable to rule out the possibility that users
% build up a mountain of type errors and then slowly reduce them to zero.

@@

Our approach to data analysis is inspired by Marceau
et~al.~\cite{mfk-onward-2011,mfk-sigcse-2011} and
Macedo et~al.~\cite{mcpcsprs-abz-2020,mcpcsprs-scp-2021}
for Racket and Alloy4Fun, respectively. Other more recent work has
also studied edit sequences to infer intent~\cite{wk-koli-2020,lgk-pj-2022,rsgl-cpp-2020}.
A major difference is that we have only samples, not full sequences,
which significantly impacts our study methods.\todo{Reasonable rewrite?}

The BitFit project collects data about beginning students interactions in
Java~\cite{ekc-wccce-2016,anna-russo-kennedy-ms-2006}; it does
not collect source code.\todo{Does it sample? If not, say so!}
However, its data do not contain details about errors.\todo{Really?!?}
Other logging-based studies of Java programs use the full error message to
draw conclusions~\cite{bgimgm-cse-2016,dlc-iticse-2014}, which we
cannot use to preserve privacy.

Much work~\cite{empirical-types,heeren-thesis} on the benefits of static types focuses on
small case studies~\cite{w-popl-1986,hw-scp-2004,td-tosem-2001},
inverviews~\cite{cdhhjklwya-hatra-2020,gstf-hatra-2021,cams-oopsla-2020},
or static
corpuses~\cite{rppd-fse-2014,bhmvv-toplas-2019,bmvv-arxiv-2019}.\todo{Why
  are we citing arXiv?}
Controlled experiments are rare;
Kleinschmager, et~al.\cite{khrts-icpc-2012} is a notable exception.
Lerner et~al.~\cite{lfgc-pldi-2007} apply a type error repair tool on
over two thousand programs.
Seidel et~al.~\cite{sscwj-oopsla-2017,sjw-jfp-2018} use five thousand programs
to train data-driven methods for localizing errors.
These are all on a vastly smaller scale than our study and also study
different issues.

>>>>

% TODO BEFORE SUBMISSION
% \usepackage[sort]{natbib} or otherwise manually fix cites to be in ascending order
% redo fig 5 to have better resolution

@@

Creators program using the
{Luau} programming language~\cite{luau-lang.org},
an extension of {Lua~5.1~\cite{lua}}.
The main addition in Luau is a static type system that infers
types for all Luau programs on the fly, as creators modify the code.
These types are used primarily in \IDE{} tooling such as autocomplete and
in API documentation~\cite{luau-autocomplete}, and indeed, creators
may be unaware that a typechecker is analyzing their code.
However, creators can opt in to receiving type error reports and they can write
their own types to guide designs and document their intentions.

Due to the broad community of creators, the goals of the Luau type
system are rather unique~\cite{bfj-hatra-2021}.
Whereas a traditional type system focuses on compilation and memory
safety, Luau takes a lenient approach by default and lets creators
gradually~\cite{st-sfp-2006,tf-popl-2008} migrate to rigorous checks
one module at a time.
In this way, for example,
students in code camps may begin in the \mnocheck{} mode that reports only
syntax errors; hobbyists can use \mnonstrict{} mode to identify
would-be runtime errors; and developers who need to ensure software
quality can rely on \mstrict{} mode.
(No matter the mode, a {\FS{}} type analysis still drives the \IDE{} tools.)
Luau's types aim to support untyped designs, in the spirit of migratory
typing~\cite{tfffgksst-snapl-2017}, so that creators can switch modes without
needing to restructure their code in major, potentially-breaking, ways.

@@

{Roblox Studio} is a workbench that
combines {3D creation} tools and an Integrated
Developer Environment (\IDE{}), as seen in \cref{fig:roblox-studio}.
The \IDE{} includes an optional \emph{Script Analysis} widget that
reports syntax errors, type errors, and problems identified by
lint tools. The main editor widget can also highlight
source locations in code where reported errors occur.

@@

\begin{figure}[t]\centering
  \includegraphics[width=.35\textwidth]{img/roblox-studio.png}
  \hspace{1cm}
  \includegraphics[width=.35\textwidth]{img/roblox-studio-ide.png}
  \caption{{Roblox Studio 3D creation} tools (left) and \IDE{} (right).}
  \label{fig:roblox-studio}
\end{figure}

@@

In addition to the main type analysis, Roblox Studio takes a second pass over
every codebase with a \emph{\FS{}} analysis to infer autocomplete suggestions
and drive other \IDE{} tools.
To infer precise types,
the \FS{} analysis uses rigorous checks very similar to those of
\mstrict{} mode (but not identical: see \cref{s:strict-vs-forcedstrict})
no matter what mode the creator declared for the script.
(Internally to Roblox Studio, this analysis is called \emph{forced strict}
because it ignores the creator's choice and applies strong checks.)
When it fails to infer a type, \FS{} analysis defaults to the unknown gradual
type~\cite{st-sfp-2006}.
Creators cannot see the errors reported by \FS{} analysis.

@@

Roblox Studio has a framework for reporting telemetry.
Various features in the \IDE{} use telemetry to measure their effectiveness.
The Luau open source toolchain does \emph{not} report telemetry,
as it is designed to be used in build environments such as Continuous Integration
servers, where hermetic deterministic builds are expected.

@@

\begin{figure}[t]\centering
  \begin{tabular}{|l|l|l|l|l|l|l|}\hline
    \uid{} & Time & Mode & Reason & Size Info & Overall Counts & Edit Range Counts
    \\\toprule
    \multicolumn{6}{c}{\raisebox{.5\normalbaselineskip}[0pt][0pt]{$\underbrace{\hspace{9cm}}$}} &
    \multicolumn{1}{c}{\raisebox{.5\normalbaselineskip}[0pt][0pt]{$\underbrace{\hspace{3cm}}$}} \\[-0.5ex]
    \multicolumn{6}{c}{fixed-size} & \multicolumn{1}{c}{variable-size}
  \end{tabular}
  \caption{Structure of a telemetry record. There are 13 overall counts and up to 70 edit range counts in each record.}
  \label{f:tdata}
\end{figure}

@@

\Cref{f:tdata} shows the structure of our telemetry records:
\begin{enumerate}
  \item
    \uid{}: a \panon{} identifier for the current session.
  \item
    Time: one timestamp from the client and one from the server.
    %% timestamp explanation \cref{s:data-cleaning}
  \item
    Mode: type analysis mode of the current file: \mnocheck{},
    \mnonstrict{}, or \mstrict{}.
  \item
    Reason: a flag that explains whether this record was sent due to a
    randomly-selected keystroke or a module switch.
  \item
    Size Info: number of lines in the codebase and number of lines in the edit range.
  \item
    Overall Counts: summarizes type errors and \FS{} errors during the
    last two invocations of type analysis, and additionally reports the number
    of times type analysis hit an internal limit on problem
    size~(see~\cref{s:code-too-complex} for details).
    For each invocation and each kind of
    analysis error, there are three summary counts:
    %% grand total = 12
    \begin{itemize}
      \item total number of errors,
      \item errors in the current module, and
      \item errors in the current edit range.
    \end{itemize}
  \item
    Edit Range Counts: these list specific errors that arose in the last two
    type analyses and overlap with the current edit range.
    Since there are 35 possible errors~(\cref{s:context}), there can be up to 70 counts
    in a telemetry record.
    Exactly how to interpret this data depends on which analysis it came from:
    \begin{itemize}
      \item
        Errors in the latest type analysis are currently visible to the
        creator.
        These may have been introduced by the changes in the edit range.
      \item
        Errors from the previous type analysis were visible before the creator
        made the latest edits.
        These may have motivated some of the changes in the edit range.
    \end{itemize}
\end{enumerate}

@@

Luau has an efficient type checker that can
analyze thousands of lines of code~(\cref{t:codebase-size}).
But, few sessions use type analysis directly and the data presents little
evidence that doing so would improve creators' quality of life.
We conjecture that Luau needs further tailoring to match
common idioms in untyped code.
Discovering these idioms is a question for future work~(\cref{s:conclusion}).

@@

Third, session-based data has limitations.
A creator who closes and reopens Roblox Studio every day has
a much higher chance of generating telemetry than one who
leaves Studio open all week.
We have no way to reliably check whether multiple sessions are
reporting on the same codebase.
Sessions may end abruptly, for instance because of machine reboot,
instead of ending when the creator was finished editing.

@@

Our telemetry has two distinguishing characteristics:
it includes no private information (both PII and source code details),
and it sends data on
randomly-selected runs of the type checker rather than specific events of interest.
To the best of our knowledge, this design occupies a unique position relative
to prior work, the two main strands of which we describe below and summarize
in~\cref{t:telemetry-design}.

In a \emph{classic} telemetry system, such as the ones in
VSCode~\cite{vscode-telemetry}, IntelliJ~\cite{intellij-telemetry}, and
\code{.NET}~\cite{dotnet-telemetry},
telemetry may include all sorts of
data: error reports, source code fragments, time\-stamps, and filepaths.
Users may be able to opt out of some telemetry, but the details depend on the
license agreement.
Furthermore, at least in VSCode, \IDE{} extensions can report their own telemetry.
While this data can be invaluable for discovering bugs in production, it
must be handled with extreme care.
The Blackbox project similarly collects telemetry in broad strokes.
Every Java program submitted for compilation goes into the Blackbox database (minus
any comments above the main class) along with the compiler's response~\cite{bkmu-sigcse-2014}.
The millions of programs in the Blackbox dataset
have enabled dozens of contributions to the area of CS
education~\cite{bask-icer-2018}, but its telemetry does little to respect privacy.

@@

At the opposite end of the privacy spectrum, Go's design for \emph{transparent}
telemetry reports only counter values~\cite{transparent-telemetry}.
Unlike Roblox Studio, transparent telemetry includes no timestamps and no 
session \uid{}s.
While useful for learning about the frequency and distribution of
specific events, the lack of timestamps and \uid{}s means it cannot track
edits over time (which we use in \cref{f:error-density}).
% BEN: commented for space.
% it cannot track 
% The data is useful for learning about the frequency of specific events and the
% distribution of events across clients (e.g., how many \code{CodeTooComplex}
% errors arise? what is the average number of type errors?).
% In~\cref{f:error-density}, for example, we see that increases to the number
% of type errors are quickly followed by decreases.
% Without timestamps, we would be unable to rule out the possibility that users
% build up a mountain of type errors and then slowly reduce them to zero.

@@

Our approach to data analysis is inspired by Marceau
et~al.~\cite{mfk-onward-2011,mfk-sigcse-2011} and
Macedo et~al.~\cite{mcpcsprs-abz-2020,mcpcsprs-scp-2021}
for Racket and Alloy4Fun, respectively. Other more recent work has
also studied edit sequences to infer intent~\cite{wk-koli-2020,lgk-pj-2022,rsgl-cpp-2020}.
A major difference is that we have only samples, not full sequences,
which significantly impacts our study methods.

The BitFit project collects data about beginning students interactions in
Java~\cite{ekc-wccce-2016,anna-russo-kennedy-ms-2006}.
It does not collect source code, nor does it collect error messages;
the telemetry simply counts all (i.e., without sampling) compile events, hint
requests, etc. in a transparent~\cite{transparent-telemetry} manner.
Other logging-based studies of Java programs use the full error message to
draw conclusions~\cite{bgimgm-cse-2016,dlc-iticse-2014}, which we
cannot use to preserve privacy.

Much work~\cite{empirical-types,heeren-thesis} on the benefits of static types focuses on
small case studies~\cite{w-popl-1986,hw-scp-2004,td-tosem-2001},
inverviews~\cite{cdhhjklwya-hatra-2020,gstf-hatra-2021,cams-oopsla-2020},
or static
corpuses~\cite{rppd-fse-2014,bhmvv-toplas-2019}
Controlled experiments are rare;
Kleinschmager, et~al.\cite{khrts-icpc-2012} is a notable exception.
Lerner et~al.~\cite{lfgc-pldi-2007} apply a type error repair tool on
over two thousand programs.
Seidel et~al.~\cite{sscwj-oopsla-2017,sjw-jfp-2018} use five thousand programs
to train data-driven methods for localizing errors.
These are all on a vastly smaller scale than our study and also study
different issues.
